% From Correlations to Photons: Relational Quantum Programming
% QPL 2026 Submission
%
% Target: 5-12 pages, EPTCS style
% Deadlines: Registration Feb 28, Paper March 7

\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{QPL 2026}

% Packages
\usepackage{amsmath,amssymb}
\usepackage{braket}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric}

\newtheorem{definition}{Definition}

% Code listing style
\lstdefinestyle{qrl}{
    language=Python,
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue},
    commentstyle=\color{gray},
    stringstyle=\color{red!70!black},
    showstringspaces=false,
    breaklines=true,
    frame=single,
    xleftmargin=2em,
    framexleftmargin=1.5em
}
\lstset{style=qrl}

% Custom commands
\newcommand{\qrl}{\textsc{QRL}}
\newcommand{\mbqc}{\textsc{MBQC}}

% Title
\title{From Correlations to Photons:\\Relational Quantum Programming}

\author{David Coldeira
\institute{Independent Researcher}
\email{dcoldeira@gmail.com}}

\def\titlerunning{From Correlations to Photons}
\def\authorrunning{D. Coldeira}

\emergencystretch=1em
\begin{document}
\maketitle

% ============================================================
% ABSTRACT
% ============================================================

\begin{abstract}
Quantum programming languages typically model computation as sequences of gate operations,
inheriting the circuit paradigm from classical computing. Yet quantum mechanics is
fundamentally a theory of correlations between systems, not operations upon them.
We explore whether this relational structure can serve as the foundation for quantum
programming. We present \qrl{} (Quantum Relational Language), a formalism where
quantum correlations are first-class primitives and measurement is modeled as
contextual inquiry. To test this approach, we express Bell and GHZ correlations
relationally and validate against known physics: CHSH parameter $S = 2.83$
(violating the classical bound of 2) and Mermin parameter $M = 4.0$
(classical limit 2). We compile \qrl{} programs to measurement-based quantum
computing (MBQC) patterns and execute them on Quandela's photonic
platform via Perceval, verified on both cloud simulator and real quantum
hardware (\texttt{qpu:belenos}). The results suggest that relational quantum
programming is viable and provides a direct compilation path to correlation-native
quantum devices.
\end{abstract}

% ============================================================
% 1. INTRODUCTION
% ============================================================

\section{Introduction}
\label{sec:introduction}

Quantum mechanics, at its core, is a theory about correlations. The Bell
inequalities~\cite{bell1964} reveal that quantum systems exhibit correlations
stronger than any classical mechanism can produce. The GHZ paradox~\cite{ghz1989,mermin1990}
demonstrates that these correlations lead to logical contradictions with local
realism---not statistically, but with certainty. Entanglement, the resource
underlying quantum computation's power, is itself a statement about correlations
between subsystems rather than properties of individual systems.

Yet the dominant paradigm for quantum programming tells a different
story. Languages like Qiskit~\cite{qiskit}, Cirq~\cite{cirq},
and Q\#~\cite{qsharp} model computation as unitary gate sequences
on qubits, a framing inherited from classical circuit design.
In this view, a Bell state is not a correlation but the
\emph{result} of applying a Hadamard followed by a CNOT:

\begin{lstlisting}
# Gate-based: Bell state as operation sequence
qc = QuantumCircuit(2)
qc.h(0)       # Hadamard on qubit 0
qc.cx(0, 1)   # CNOT from 0 to 1
\end{lstlisting}

This gate-centric paradigm has proven effective, but it raises a question:
\emph{what if we programmed the correlations directly?}

\subsection{The Relational Perspective}

We explore an alternative approach grounded in the relational structure of
quantum mechanics. Rather than constructing correlations from gate sequences,
we treat correlations as primitive programming constructs. A Bell pair is not
built---it is declared:

\begin{lstlisting}
# Relational: Bell state as correlation declaration
bell = program.entangle(alice, bob)  # These systems are correlated
\end{lstlisting}

This shift in perspective aligns with foundational work on relational quantum
mechanics~\cite{rovelli1996}, which argues that quantum states describe
relations between systems rather than intrinsic properties. It also connects
naturally to measurement-based quantum computing (\mbqc{})~\cite{raussendorf2001},
where computation proceeds through measurements on entangled resource states
rather than gate application.

We do not position the relational approach as a replacement for
gate-based programming---comparative evaluation is future work. What we
demonstrate is that it is \emph{viable}: correlations can serve as
first-class programming primitives, and programs written this way
compile to photonic platforms and produce correct physics.

\subsection{Contributions}

We present \qrl{} (Quantum Relational Language), a formalism for exploring
relational quantum programming, and report on our investigation:

\begin{enumerate}
    \item \textbf{A relational formalism} (Section~\ref{sec:formalism}):
    \texttt{QuantumRelation} as a primitive for correlated systems;
    measurement as contextual inquiry.

    \item \textbf{Quantum correlation tests} (Section~\ref{sec:correlations}):
    Bell (CHSH) and GHZ (Mermin) in \qrl{}, validated against known physics:
    $S = 2.83$ (Tsirelson bound), $M = 4.0$ (classical limit: 2).

    \item \textbf{MBQC compilation} (Section~\ref{sec:mbqc}):
    \qrl{} relations compile to \mbqc{} patterns with automatic graph
    extraction and adaptive corrections (teleportation fidelity $F=1.0$).

    \item \textbf{Photonic execution} (Section~\ref{sec:photonic}):
    End-to-end pipeline via Perceval~\cite{perceval} to Quandela's platform,
    verified on \texttt{sim:belenos} and \texttt{qpu:belenos}.
\end{enumerate}

Our implementation comprises approximately 6,300 lines of Python with 206
tests. The code is available at \url{https://github.com/dcoldeira/quantum-relational-language}.

Our contribution is integrating these components through a relational
lens: \qrl{} maps relations directly to graph states, bypassing the
circuit decomposition step that gate-based \mbqc{} targets require.
Whether this yields practical efficiency gains remains to be benchmarked.
We claim relational quantum programming \emph{works}: correct physics,
hardware execution, cross-platform portability.

% ============================================================
% 2. THE RELATIONAL FORMALISM
% ============================================================

\section{The Relational Formalism}
\label{sec:formalism}

We now present the \qrl{} formalism. Our goal is not to define a complete
programming language with formal type theory, but to introduce the core
abstractions that enable relational quantum programming and establish
their semantics precisely enough to support the results that follow.

\subsection{Design Philosophy}

\qrl{} is built on three principles drawn from the foundations of quantum mechanics:

\paragraph{Relations over objects.}
In Rovelli's relational quantum mechanics~\cite{rovelli1996}, quantum states
describe relations between systems, not intrinsic properties.
\qrl{} operationalizes this view: the primitive construct is a
\texttt{QuantumRelation} representing correlated systems, not individual
qubits with independent states.

\paragraph{Correlations as primitive.}
In \qrl{}, entanglement is declared directly rather than derived from gate
sequences: the Bell correlation \emph{is} the program. This reflects a
different design philosophy---\emph{what} correlations exist, not \emph{how}
to construct them.

\paragraph{Measurement as inquiry.}
\qrl{} models measurement as \emph{asking a question} of a correlated system,
with explicit specification of basis and observer context, connecting to
quantum contextuality~\cite{coecke2017}: the answer depends not just on the
system but on the question asked.

\subsection{Core Constructs}

\qrl{} introduces three core abstractions:

\begin{definition}[Quantum Relation]
A \texttt{QuantumRelation} is a tuple $R = (\mathcal{S}, \ket{\psi}, S)$ where:
\begin{itemize}
    \item $\mathcal{S} = \{s_1, \ldots, s_n\}$ is a set of quantum system identifiers
    \item $\ket{\psi} \in \mathcal{H}^{\otimes n}$ is the joint quantum state
    \item $S \in [0, \log_2 d]$ is the entanglement entropy (von Neumann entropy
          of the reduced density matrix), where $d$ is the dimension of the
          smaller subsystem in a bipartition. For a single qubit reduced state,
          $S \in [0, 1]$.
\end{itemize}
\end{definition}

The key distinction from gate-based models: a \texttt{QuantumRelation} represents
the \emph{correlation structure} between systems, not a state that will be
operated upon. The systems in $\mathcal{S}$ are correlated by virtue of being
in the relation; their joint state $\ket{\psi}$ encodes the specific correlation.

\begin{definition}[Quantum Question]
A \texttt{QuantumQuestion} is a tuple $Q = (B, i, t)$ where:
\begin{itemize}
    \item $B \in U(2)$ is the measurement basis (unitary defining the eigenbasis)
    \item $i \in \{1, \ldots, n\}$ is the target subsystem index
    \item $t \in \{\text{SPIN\_Z}, \text{SPIN\_X}, \text{SPIN\_Y}, \text{CUSTOM}\}$
          is the question type
\end{itemize}
\end{definition}

Measurement in \qrl{} is performed by the \texttt{ask} operation, which takes
a relation and a question:

\begin{lstlisting}
result = program.ask(relation, question, perspective="alice")
\end{lstlisting}

The optional \texttt{perspective} parameter annotates the observer
context, aligning with relational QM~\cite{rovelli1996}; extending it
to affect state update scope in distributed protocols is future work.

\subsection{Operational Semantics}

We define the semantics of \qrl{}'s core operations.

\paragraph{Entanglement.}
The \texttt{entangle} operation creates a \texttt{QuantumRelation} from $n$
quantum systems. For GHZ-type entanglement (the default):
\begin{equation}
\texttt{entangle}(s_1, \ldots, s_n) \mapsto R = (\{s_1, \ldots, s_n\}, \ket{GHZ_n}, 1)
\end{equation}
where $\ket{GHZ_n} = \frac{1}{\sqrt{2}}(\ket{0}^{\otimes n} + \ket{1}^{\otimes n})$
and the entanglement entropy is maximal ($S = 1$) for any bipartition.

For Bell states specifically ($n = 2$):
\begin{equation}
\texttt{entangle}(s_1, s_2) \mapsto R = (\{s_1, s_2\}, \ket{\Phi^+}, 1)
\end{equation}
where $\ket{\Phi^+} = \frac{1}{\sqrt{2}}(\ket{00} + \ket{11})$.

\paragraph{Measurement.}
The \texttt{ask} operation implements standard projective measurement
(Born rule). Given relation $R = (\mathcal{S}, \ket{\psi}, S)$ and
question $Q = (B, i, t)$: transform to the measurement basis via
$B^\dagger$ on subsystem $i$, sample outcome $m$ with probability
$P(m) = \sum_{j: j_i = m} |\psi'_j|^2$, then project and renormalize.
The semantics are standard quantum measurement framed as ``asking a
question''---the physics is identical, but the conceptual framing
emphasizes the relational structure.

\subsection{Example: Bell Correlations}

To illustrate, consider expressing Bell correlations in \qrl{}:

\begin{lstlisting}
bell = program.entangle(alice, bob)      # Declare correlation
a = program.ask(bell, SPIN_Z, subsystem=0, perspective="alice")
b = program.ask(bell, SPIN_Z, subsystem=1, perspective="bob")
# Outcomes are perfectly correlated: a == b always
\end{lstlisting}

The correlation is inherent in the \texttt{QuantumRelation}, not computed
from gate sequences. We did not \emph{construct} a Bell state and then
\emph{measure} it---we \emph{declared} that Alice and Bob share Bell
correlations, then \emph{asked questions} about those correlations.

% ============================================================
% 3. MODELING QUANTUM CORRELATIONS
% ============================================================

\section{Modeling Quantum Correlations}
\label{sec:correlations}

Having defined the \qrl{} formalism, we now test whether it can express
fundamental quantum correlations and reproduce known physics. We implement
two canonical tests: the Bell-CHSH inequality and the GHZ-Mermin inequality.
These are not arbitrary benchmarks---they represent the sharpest demonstrations
that quantum correlations violate classical expectations.

\subsection{Bell Test: CHSH Inequality}

The CHSH inequality~\cite{chsh1969} provides a quantitative bound on correlations
achievable by any local hidden-variable theory. Consider two parties, Alice and
Bob, sharing a correlated system. Alice chooses to measure in basis $a$ or $a'$;
Bob chooses $b$ or $b'$. The correlation function is the expectation value
of the product of $\pm 1$ outcomes~\cite{nielsen2010}:
\begin{equation}
E(x, y) = \langle A_x B_y \rangle = P(\text{same}|x,y) - P(\text{different}|x,y)
\end{equation}
where $x \in \{a, a'\}$ and $y \in \{b, b'\}$. The CHSH parameter is:
\begin{equation}
S = E(a,b) - E(a,b') + E(a',b) + E(a',b')
\end{equation}

\paragraph{Classical bound.}
Any local hidden-variable theory satisfies $|S| \leq 2$.

\paragraph{Quantum bound.}
Quantum mechanics permits $|S| \leq 2\sqrt{2} \approx 2.83$ (the Tsirelson
bound~\cite{tsirelson1980}), achieved by measuring a Bell state $\ket{\Phi^+}$
with optimal angles: $a = 0$, $a' = \pi/2$, $b = \pi/4$, $b' = 3\pi/4$.

\paragraph{QRL implementation.}
In \qrl{}, we express the Bell test as follows:

\begin{lstlisting}
# Declare Bell correlations
bell = program.entangle(alice, bob)

# Compute CHSH parameter
S = chsh_parameter(bell,
    alice_angles=[0, np.pi/2],
    bob_angles=[np.pi/4, 3*np.pi/4],
    shots=10000)
\end{lstlisting}

The \texttt{chsh\_parameter} function computes correlations by repeatedly
asking questions of the Bell relation at the specified measurement angles,
then combining results according to the CHSH formula.

\paragraph{Results.}
\qrl{} produces $S = 2\sqrt{2} \approx 2.83$, saturating the Tsirelson
bound. The value can be computed analytically from the correlation function
$E(a,b) = \cos(a - b)$ for $\ket{\Phi^+}$, and is confirmed by Monte Carlo
simulation (10{,}000 measurement trials per setting). The classical bound
of 2 is decisively violated.

\begin{table}[ht]
\centering
\begin{tabular}{lcc}
\toprule
\textbf{Bound} & \textbf{Value} & \textbf{QRL Result} \\
\midrule
Classical (local hidden variables) & $|S| \leq 2$ & --- \\
Quantum (Tsirelson bound~\cite{tsirelson1980}) & $|S| \leq 2\sqrt{2}$ & $S = 2.83$ \\
\bottomrule
\end{tabular}
\caption{CHSH inequality results. \qrl{} achieves the quantum maximum.}
\label{tab:chsh}
\end{table}

The key observation: we did not ``program'' a Bell test by constructing
circuits. We declared correlations and asked questions; the violation
emerged from the relational structure.

\subsection{GHZ Paradox: Mermin Inequality}

The GHZ paradox~\cite{ghz1989} provides an even starker demonstration of
quantum non-classicality. Unlike Bell inequalities, which show statistical
violations, the GHZ paradox demonstrates a logical contradiction with
local realism---a ``proof without inequalities.''

\paragraph{Setup.}
Three parties (Alice, Bob, Charlie) share a GHZ state:
\begin{equation}
\ket{GHZ_3} = \frac{1}{\sqrt{2}}(\ket{000} + \ket{111})
\end{equation}

Consider measurements in the $X$ and $Y$ bases. Quantum mechanics predicts:
\begin{itemize}
    \item $XXX$ measurement: product of outcomes is always $+1$
    \item $XYY$ measurement: product of outcomes is always $-1$
    \item $YXY$ measurement: product of outcomes is always $-1$
    \item $YYX$ measurement: product of outcomes is always $-1$
\end{itemize}

\paragraph{The paradox.}
If outcomes were determined by local hidden variables, multiplying the four
constraint equations yields a contradiction: the product of all predetermined
values appears squared (hence $+1$), but the right-hand side is
$(+1)(-1)(-1)(-1) = -1$. No local hidden-variable assignment can satisfy all
four correlations simultaneously~\cite{mermin1990}.

\paragraph{Mermin inequality.}
The Mermin inequality~\cite{mermin1990} quantifies this. Define:
\begin{equation}
M = \langle XXX \rangle - \langle XYY \rangle - \langle YXY \rangle - \langle YYX \rangle
\end{equation}

Classical bound: $|M| \leq 2$. Quantum prediction for GHZ: $M = 4$.

\paragraph{QRL implementation.}
The GHZ test follows the same pattern:

\begin{lstlisting}
# Declare GHZ correlations
ghz = program.entangle(alice, bob, charlie, state_type="ghz")

# Test the paradox
xxx = ghz_correlation(ghz, bases=['X', 'X', 'X'])  # Always +1
xyy = ghz_correlation(ghz, bases=['X', 'Y', 'Y'])  # Always -1

# Compute Mermin parameter
M = mermin_parameter_3(ghz, shots=10000)
\end{lstlisting}

\paragraph{Results.}

\begin{table}[ht]
\centering
\begin{tabular}{lcc}
\toprule
\textbf{Measurement} & \textbf{QM Prediction} & \textbf{QRL Result} \\
\midrule
$XXX$ eigenvalue & $+1$ & $+1$ \\
$XYY$ eigenvalue & $-1$ & $-1$ \\
$YXY$ eigenvalue & $-1$ & $-1$ \\
$YYX$ eigenvalue & $-1$ & $-1$ \\
\midrule
Mermin $M$ (classical limit: 2) & $4.0$ & $4.0$ \\
\bottomrule
\end{tabular}
\caption{GHZ-Mermin results. \qrl{} reproduces quantum predictions exactly.}
\label{tab:mermin}
\end{table}

The eigenvalues are exact (not statistical) because $\ket{GHZ_3}$ is a
simultaneous eigenstate of the operators $XXX$, $XYY$, $YXY$, and $YYX$:
\qrl{} produces $+1$ for every $XXX$ trial and $-1$ for every $XYY$, $YXY$,
$YYX$ trial. The Mermin parameter $M = 4.0$ maximally violates the classical
bound of 2.

\subsection{Discussion}

These results demonstrate that \qrl{}'s relational formalism correctly captures
quantum correlations. The CHSH violation ($S = 2.83$) and Mermin violation
($M = 4.0$) match theoretical predictions precisely.

\paragraph{The relational insight.}
The relational approach---declaring correlations as primitives rather than
constructing them from gates---produces physically correct results. A
gate-based implementation of these same tests would produce identical
physics; the value of the relational framing lies not in the results
themselves but in what the formalism makes explicit. We did not ``run
an algorithm'' on a Bell state; we declared that Alice and Bob share
Bell correlations and asked what questions would reveal. The violation
is not computed---it is inherent in the relational structure we specified.

Whether this framing offers practical advantages for reasoning about
quantum protocols remains an open question (Section~\ref{sec:conclusion}).

% ============================================================
% 4. COMPILATION TO MBQC
% ============================================================

\section{Compilation to MBQC}
\label{sec:mbqc}

The previous section demonstrated that \qrl{} correctly models quantum
correlations. But a programming formalism must also \emph{execute}. We now
show that \qrl{} programs compile naturally to measurement-based quantum
computing (\mbqc{}) patterns, providing a path to hardware execution.

\subsection{Why MBQC?}

In \mbqc{}~\cite{raussendorf2001,raussendorf2003}, computation proceeds by
preparing an entangled resource state (graph state) and performing adaptive
single-qubit measurements. \mbqc{} is relevant for two reasons: it is the
native model for photonic hardware~\cite{rudolph2017} (where generating
entangled photons is easier than applying deterministic gates), and \qrl{}'s
\texttt{QuantumRelation} abstraction directly encodes the entanglement
structure that \mbqc{} requires. The compilation is structural, not
transformational: since \qrl{} relations directly encode graph state
topology, the extraction preserves entanglement structure by
construction. The subsequent \mbqc{} pattern generation and adaptive
corrections follow the standard universality proof of
Raussendorf et al.~\cite{raussendorf2003}, which establishes that any
unitary can be realized by single-qubit measurements on a sufficiently
large graph state.

\subsection{The Compilation Pipeline}

\qrl{} compilation proceeds in three stages (Figure~\ref{fig:pipeline}).
The first three stages are backend-independent; only the final encoding
step is platform-specific.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    node distance=0.7cm and 1.1cm,
    box/.style={draw, rounded corners, minimum height=0.7cm,
                minimum width=1.4cm, align=center, font=\small},
    backend/.style={draw, rounded corners, minimum height=0.7cm,
                minimum width=1.2cm, align=center, font=\scriptsize},
    arr/.style={-{Stealth[length=2mm]}, thick}
]
\node[box, fill=blue!10] (qrl) {\qrl{} Program};
\node[box, fill=green!10, right=of qrl] (graph) {Graph State};
\node[box, fill=orange!10, right=of graph] (mbqc) {\mbqc{} Pattern};
\node[backend, fill=red!10, above right=0.4cm and 0.9cm of mbqc] (perc) {Perceval};
\node[backend, fill=purple!10, right=of perc] (cloud) {Quandela};
\node[backend, fill=red!10, below right=0.4cm and 0.9cm of mbqc] (penny) {PennyLane};
\node[backend, fill=purple!10, right=of penny] (sim) {Simulation};
\draw[arr] (qrl) -- node[above, font=\scriptsize] {extract} (graph);
\draw[arr] (graph) -- node[above, font=\scriptsize] {compile} (mbqc);
\draw[arr] (mbqc.east) -- (perc.west);
\draw[arr] (perc) -- (cloud);
\draw[arr] (mbqc.east) -- (penny.west);
\draw[arr] (penny) -- (sim);
\end{tikzpicture}
\caption{The \qrl{} execution pipeline: from relational specification
through \mbqc{} compilation to backend-specific execution. The core
pipeline (left) is platform-independent.}
\label{fig:pipeline}
\end{figure}

\paragraph{Stage 1: Graph extraction.}
Every \texttt{QuantumRelation} corresponds to a graph state. The systems in
the relation become vertices; the entanglement structure determines edges.

\begin{itemize}
    \item \textbf{Bell relation} $\rightarrow$ 2-vertex graph with single edge
    \item \textbf{GHZ relation} $\rightarrow$ star graph with local Clifford corrections
    \item \textbf{Linear cluster} $\rightarrow$ path graph
\end{itemize}

Formally, for a relation $R = (\mathcal{S}, \ket{\psi}, S)$, we extract
graph $G = (V, E)$ where $V = \mathcal{S}$ and $E$ encodes the entanglement
connectivity (Figure~\ref{fig:graphs}). For GHZ-type states, the star graph
produces a graph state that is local-Clifford equivalent to the GHZ state,
not identical to it~\cite{raussendorf2003}. The compiler applies the
necessary single-qubit corrections to recover the target state.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    node distance=1.2cm,
    qubit/.style={circle, draw, fill=blue!20, minimum size=0.6cm,
                  font=\small},
    lbl/.style={font=\small\itshape, text=gray}
]
% Bell relation
\node[lbl] at (0, 1.2) {Bell relation};
\node[qubit] (b1) at (-0.6, 0) {$s_1$};
\node[qubit] (b2) at (0.6, 0) {$s_2$};
\draw[thick] (b1) -- (b2);

% GHZ-3 relation
\node[lbl] at (3.5, 1.2) {GHZ-3 relation};
\node[qubit] (g1) at (3.5, 0) {$s_1$};
\node[qubit] (g2) at (2.3, -0.9) {$s_2$};
\node[qubit] (g3) at (4.7, -0.9) {$s_3$};
\draw[thick] (g1) -- (g2);
\draw[thick] (g1) -- (g3);

% Linear cluster
\node[lbl] at (7.5, 1.2) {Linear cluster};
\node[qubit] (l1) at (6.3, 0) {$s_1$};
\node[qubit] (l2) at (7.5, 0) {$s_2$};
\node[qubit] (l3) at (8.7, 0) {$s_3$};
\draw[thick] (l1) -- (l2);
\draw[thick] (l2) -- (l3);
\end{tikzpicture}
\caption{Graph states extracted from \qrl{} relations. Bell: single edge.
GHZ: star graph (with local Clifford corrections). Cluster: path graph.}
\label{fig:graphs}
\end{figure}

\paragraph{Stage 2: Pattern generation.}
Each \texttt{ask} operation in the \qrl{} program becomes a measurement in
the \mbqc{} pattern. The measurement basis in \qrl{} determines the
measurement angle in \mbqc{}:

\begin{itemize}
    \item $Z$-basis measurement $\rightarrow$ angle $\theta = 0$
    \item $X$-basis measurement $\rightarrow$ angle $\theta = \pi/2$
    \item Arbitrary basis $R_z(\phi)$ $\rightarrow$ angle $\theta = \phi$
\end{itemize}

The pattern also tracks measurement order and dependencies for adaptive
corrections.

\paragraph{Stage 3: Adaptive corrections.}
\mbqc{} requires classical feedforward: later measurements may need angle
adjustments based on earlier outcomes. \qrl{} automatically computes these
corrections using the standard \mbqc{} Pauli frame tracking rules.

For a measurement $M_j$ depending on prior outcome $m_i$:
\begin{equation}
\theta_j' = (-1)^{m_i \cdot s_{ij}} \theta_j + m_i \cdot r_{ij} \pi
\end{equation}
where $s_{ij}$ and $r_{ij}$ encode the signal and correction dependencies
from the graph structure.

\subsection{Example: Bell State Compilation}

Consider a simple \qrl{} program:

\begin{lstlisting}
bell = program.entangle(q0, q1)
m0 = program.ask(bell, SPIN_Z, subsystem=0)
m1 = program.ask(bell, SPIN_X, subsystem=1)
\end{lstlisting}

This compiles to graph $G = (\{q_0, q_1\}, \{(q_0, q_1)\})$---the minimal
cluster state for a Bell pair. The \mbqc{} pattern prepares $\ket{+}\ket{+}$,
applies CZ, then measures $q_0$ at angle 0 and $q_1$ at angle $\pi/2$, with
a $Z$ correction on the output conditioned on $m_0 = 1$.

\subsection{Teleportation Validation}

To validate the pipeline, we implemented quantum teleportation---a protocol
requiring correct \mbqc{} patterns and adaptive corrections. The sender and
receiver share a Bell relation; the sender performs a joint measurement and
communicates the outcome; the receiver applies an adaptive correction.

We tested on six input states ($\ket{0}$, $\ket{1}$, $\ket{+}$, $\ket{-}$,
$\ket{i}$, and an arbitrary superposition). All achieved fidelity $F = 1.0$,
confirming correct graph extraction, pattern generation, and adaptive
correction computation.

% ============================================================
% 5. PHOTONIC EXECUTION
% ============================================================

\section{Photonic Execution}
\label{sec:photonic}

The compilation pipeline of Section~\ref{sec:mbqc} produces \mbqc{} patterns.
To complete the path ``from correlations to photons,'' we must execute these
patterns on a photonic computing platform. This section describes our integration with
Quandela's photonic platform and reports execution results.

\subsection{The Full Pipeline}

Our execution pipeline (Figure~\ref{fig:pipeline}) uses
graphix~\cite{graphix} as an \mbqc{} intermediate representation,
Perceval~\cite{perceval} for photonic circuit construction and local
simulation (SLOS backend), and Quandela's cloud platform for remote
execution. We target Quandela's Belenos platform, running on both
\texttt{sim:belenos} (cloud simulator) and \texttt{qpu:belenos}
(Quandela's 12-qubit photonic QPU).

\subsection{The Encoding Challenge}

The graphix-perceval bridge produces \emph{polarization-encoded} circuits
(PBS, wave plates), but Quandela's cloud simulator supports only
\emph{path encoding}. Submitting polarization-encoded circuits to
\texttt{sim:belenos} returned no results.

We solved this by building a direct \qrl{} $\rightarrow$ Perceval adapter
that generates path-encoded (dual-rail) circuits, where each qubit uses
two optical modes: $\ket{0} \rightarrow \ket{1,0}$ and
$\ket{1} \rightarrow \ket{0,1}$. The adapter translates Bell relations
to 4-mode circuits (2 photons), GHZ-3 to 6-mode circuits (3 photons),
and handles non-adjacent mode entanglement via permutation gates.
We validated it with 22 dedicated tests.

\subsection{Cloud Execution Results}

With the path-encoding adapter, we achieved end-to-end execution from
\qrl{} to Quandela's cloud platform.

\paragraph{Bell state.}
The complete pipeline for a Bell state:

\begin{lstlisting}
bell = program.entangle(alice, bob)       # QRL relation
circuit = qrl_to_perceval(compile_to_mbqc(bell))  # 4 modes, 2 photons
result = run_on_cloud(circuit, platform="sim:belenos", shots=1000)
\end{lstlisting}

The circuit executed successfully on \texttt{sim:belenos} with 1{,}000
shots. Bunched outputs ($\ket{2,0,0,2}$, $\ket{0,2,2,0}$) dominated at
equal rates, consistent with $\ket{\Phi^+}$ symmetry and local SLOS
simulation. We subsequently ran the same circuit on \texttt{qpu:belenos},
Quandela's 12-qubit photonic QPU. Of 1{,}000 shots, 423 yielded valid
dual-rail events (42.3\%); the remaining 577 were HOM-bunched. Among
valid events: $\ket{10}$ 33.6\%, $\ket{00}$ 24.6\%, $\ket{11}$ 24.3\%,
$\ket{01}$ 17.5\%. The approximately uniform distribution is expected for
linear optics without active feed-forward; the 57.7\% HOM fraction
is consistent with 2-photon interference on a real device.

\paragraph{GHZ-3 state.}
The GHZ-3 circuit (6 modes, 3 photons) executed on \texttt{sim:belenos}
but with low yield: 1 sample passed post-selection out of 1{,}000 shots.
This is expected---multi-photon HOM bunching suppresses valid dual-rail
events exponentially. Local SLOS simulation confirmed correct GHZ
correlations.

\subsection{Summary}

\begin{table}[ht]
\centering
\begin{tabular}{lcccc}
\toprule
\textbf{State} & \textbf{Modes} & \textbf{Photons} & \textbf{sim:belenos} & \textbf{qpu:belenos} \\
\midrule
Bell ($\ket{\Phi^+}$) & 4 & 2 & Verified & Verified (423/1000 valid) \\
GHZ-3 & 6 & 3 & Executed (low yield) & Not attempted \\
\bottomrule
\end{tabular}
\caption{Photonic execution results on Quandela cloud platforms.}
\label{tab:photonic}
\end{table}

The complete pipeline---from \qrl{} relation to \mbqc{} pattern to
path-encoded Perceval circuit---is operational on both simulator and
real hardware, producing physically meaningful results.

% ============================================================
% 6. RELATED WORK
% ============================================================

\section{Related Work}
\label{sec:related}

We position \qrl{} relative to existing work across four categories.

\paragraph{Quantum programming languages.}
Selinger's QPL~\cite{selinger2004} established foundational semantics for
quantum programming via superoperators. Quipper~\cite{quipper} is a scalable
circuit language in Haskell; QWIRE~\cite{qwire} brings linear types to
circuit construction. Dominant frameworks---Qiskit~\cite{qiskit},
Cirq~\cite{cirq}, Q\#~\cite{qsharp}---model computation as gate sequences.
\qrl{} differs in treating correlations rather than gates as primitives, and
does not yet provide formal denotational semantics; we explore whether a
relational foundation is feasible, not whether it supersedes existing approaches.

\paragraph{MBQC frameworks.}
graphix~\cite{graphix} provides simulation and compilation for \mbqc{},
representing computation as graph states with measurement patterns.
MentPy~\cite{mentpy} focuses on parametrized \mbqc{} for quantum machine learning.
Both frameworks work directly with graph states and measurement angles.
\qrl{} operates at a higher abstraction level: users declare correlations,
and the system extracts graph states automatically. We use graphix as
a compilation target, not a competitor.

\paragraph{Photonic frameworks.}
Perceval~\cite{perceval} and Strawberry Fields~\cite{strawberryfields} target
linear optical hardware at the component level (beamsplitters, phase shifters).
\qrl{} compiles \emph{from} relational specifications \emph{to} photonic
circuits, providing an abstraction layer above component-level programming.
PennyLane~\cite{pennylane} provides a hardware-agnostic interface;
\qrl{} also targets it via mid-circuit measurements, confirming
cross-platform portability.

\paragraph{Formal and categorical approaches.}
ZX-calculus~\cite{zxcalculus} and categorical quantum
mechanics~\cite{coecke2017} provide diagrammatic foundations for
quantum reasoning, connecting naturally to \mbqc{}. \qrl{}'s
relational structure may admit categorical semantics---relations
as morphisms in a suitable category---but we leave this
formalization to future work.

\paragraph{Relational quantum mechanics.}
Rovelli's relational quantum mechanics~\cite{rovelli1996} inspired
\qrl{}'s design. However, \qrl{} is a programming formalism, not a
physical interpretation---we operationalize relational thinking for
computation without committing to metaphysical claims.

\paragraph{Quantum causal models.}
Allen et al.~\cite{allen2017} show that Bell correlations have a causal
explanation without fine-tuning when common causes are quantum.
\qrl{}'s \texttt{QuantumRelation}---a first-class declared correlation,
not a circuit-derived state---is structurally analogous to such a quantum
common cause node. Morimae~\cite{morimae2014} shows that acausal \mbqc{} is resourced
by a process matrix $W \!=\! 2^{N+n}|G'\rangle\langle G'|\otimes(I/2)^{\otimes n}$;
\qrl{}'s gflow enforcement places compiled patterns in the causally
separable regime. We identified these
alignments after this work and document them in a companion research note.

% ============================================================
% 7. CONCLUSION
% ============================================================

\section{Conclusion}
\label{sec:conclusion}

We set out to explore a question: can quantum correlations serve as
first-class programming primitives? The gate-based paradigm has proven
enormously successful, but quantum mechanics is fundamentally a theory
of correlations. We wondered whether programming could reflect this
relational structure directly.

\paragraph{Summary.}
\qrl{} treats correlations as primitives: a Bell pair is declared, not
constructed; measurement is inquiry, not disturbance. The formalism
reproduces known quantum correlations exactly (CHSH $S = 2.83$, Mermin
$M = 4.0$), compiles to \mbqc{} patterns (teleportation fidelity $F = 1.0$),
and executes on Quandela's photonic platform---verified on
\texttt{sim:belenos} and the 12-qubit \texttt{qpu:belenos} QPU; a PennyLane
backend confirms cross-platform portability. Whether bypassing circuit
decomposition yields efficiency gains remains to be benchmarked.
We claim relational quantum programming is \emph{different}, \emph{viable},
and \emph{worth exploring}: the correlations that make quantum mechanics
distinctive are primitives to be programmed, not side effects to be managed.

% ============================================================
% ACKNOWLEDGMENTS
% ============================================================

\paragraph{Acknowledgments.}
We thank Quandela for providing cloud access to their photonic
computing platform, and the developers of graphix, Perceval, and PennyLane
for their open-source frameworks. We also thank the anonymous
reviewers for constructive feedback that improved this paper.

% ============================================================
% REFERENCES
% ============================================================

\bibliographystyle{eptcs}
\bibliography{references}

\end{document}
